---
title: "Literal Listener"
output: html_document
---

```{r global_options, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  echo=F, warning=F, cache=T, 
  message=F, sanitiz =T, 
  fig.width = 5, fig.height = 2)
```

```{r load_settings}
source("~/Settings/startup.R")
```

```{r}
literal = function(number, utterance) {
  program_file = paste("stories/lk", number,
                       "/autoexpanded.wppl", sep="")
  return(webppl(
    program_file = program_file,
    inference_opts = list(method="enumerate"),
    model_var = paste(
      "literal({utterance: '",
      utterance, "'})", 
      sep="")
  ) %>% mutate(utterance=utterance))
}

plot_literal = function(number, explanations) {
  rs = rbind(
    do.call(rbind,
            lapply(explanations, function(expl) {
              return(literal(number, expl))
            })),
    literal(number, "null"))
  p = rs %>%
    gather(variable, value, -prob, -utterance) %>%
    group_by(variable, utterance) %>%
    summarise(marginal.prob = sum(prob[value])) %>%
    separate(variable, c("type", "variable")) %>%
    ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
    geom_bar(stat="identity", position="dodge") +
    facet_wrap(~type, scale="free") +
    theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
    scale_fill_brewer(type="qual", palette = 2) +
    ylim(0, 1)
  
  print(p)
  
  return(rs)
}
```

```{r}
print_world = function(
  a=F, b=F, c=F, d=F,
  cause=F, dir=F,
  ab=F, bc=F, ac=F, cd=F) {
  return(paste(
    ifelse(a, "A", ""),
    ifelse(b, "B", ""),
    ifelse(c, "C", ""),
    ifelse(d, "D", ""),
    ifelse(cause|dir|ab|ac|bc|cd, "|", ""),
    ifelse(cause, "c", ""),
    ifelse(dir, "d", ""),
    ifelse(ab, "ab|", ""),
    ifelse(ac, "ac|", ""),
    ifelse(bc, "bc|", ""),
    ifelse(cd, "cd|", "")
    , sep=""))
}
```


### Story 1
```{r}
l1 = plot_literal(1, list("A because B"))
```

```{r}
l1 %>% filter(output.A & output.B) %>%
  group_by(utterance) %>%
  mutate(prob = prob/sum(prob)) %>%
  ungroup() %>%
  as.data.frame() %>%
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1) +
  ggtitle("presuppose A and B")
```

### Story 2

```{r}
l2 = plot_literal(2, c("A because B", "B because A", "C because A"))
```

Why does the marginal for `causeAB` go down given "B because A"?

```{r}
l2bcausea = literal(2, "B because A")
```

I guess B is more likely to be true if it's deterministically true, i.e. if there's no cause. So if "B" is part of the meaning, then that will make a difference.

```{r}
l2bcausea %>% mutate(world = print_world(
  a=output.A, b=output.B, c=output.C,
  ab=structureParams.causeAB, bc=structureParams.causeBC)) %>%
  filter(structureParams.causeAB)
```

If we're given that B is true, then saying "A because B" or "B because A" or "C because A" increases the probability of the causes. But I think this is only because they make A more likely.

```{r}
l2 %>% 
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1)
```

Apparently, given that B is true, causeAB is pretty unlikely a priori.

```{r}
l2 %>% filter(output.B) %>%
  group_by(utterance) %>%
  mutate(prob = prob/sum(prob)) %>%
  ungroup() %>%
  as.data.frame() %>%
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1) +
  ggtitle("presuppose B")
```

Given both A and B, "B because A" becomes the best indicator for `causeAB` and "C because A" is a good indicator for both `causeAB` and `causeBC`.

```{r}
l2 %>% filter(output.A & output.B) %>%
  group_by(utterance) %>%
  mutate(prob = prob/sum(prob)) %>%
  ungroup() %>%
  as.data.frame() %>%
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1) +
  ggtitle("presuppose A and B")
```

Assuming C, too, shows a much more intuitive interpretation of the explanation.

```{r}
l2 %>% filter(output.A & output.B & output.C) %>%
  group_by(utterance) %>%
  mutate(prob = prob/sum(prob)) %>%
  ungroup() %>%
  as.data.frame() %>%
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1) +
  ggtitle("presuppose A, B, and C")
```

### Story 3
```{r}
l3 = plot_literal(3, c("B because ! C", "! C because A", "! C because B"))
```

```{r}
l3 %>% filter(output.A & output.B & !output.C) %>%
  group_by(utterance) %>%
  mutate(prob = prob/sum(prob)) %>%
  ungroup() %>%
  as.data.frame() %>%
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1) +
  ggtitle("presuppose A, B, and !C")
```

<!--
### Story 4
```{r}
l4 = plot_literal(4, c("! C because ! A", "! D because !B", "! A because !D"))
```

```{r}
l4 %>% filter(!output.A & !output.B & !output.C & !output.D) %>%
  group_by(utterance) %>%
  mutate(prob = prob/sum(prob)) %>%
  ungroup() %>%
  as.data.frame() %>%
  gather(variable, value, -prob, -utterance) %>%
  group_by(variable, utterance) %>%
  summarise(marginal.prob = sum(prob[value])) %>%
  separate(variable, c("type", "variable")) %>%
  ggplot(., aes(x=variable, y=marginal.prob, fill=utterance)) +
  geom_bar(stat="identity", position="dodge") +
  facet_wrap(~type, scale="free") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  scale_fill_brewer(type="qual", palette = 2) +
  ylim(0, 1) +
  ggtitle("presuppose !A, !B, !C, and !D")
```
-->