---
title: "RSA Explanations"
author: "Erin Bennett"
output: 
  html_document:
      toc: false
bibliography: "../../explanations.bib"
---

```{r global_options, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo=F, warning=F, cache=T, message=F, sanitiz =T, fig.width = 5, fig.height = 3)
```

```{r load_settings}
source("~/cocolab/settings/startup.R")
library(reshape2)
```

There are several components to an explanation:

* Actual State
    * observable variables (input, output)
    * latent variables (structural parameters, unobservable states)
* Counterfactual Prior over Programs
    * structural parameters
    * alternative inputs
* Listener
    * knowledge about observable variables
    * knowledge about latent variables
* Speaker
    * alternative utterances (and their associated costs)
    * question under discussion (utility function)
    * knowledge about actual state
* Literal
    * a semantics for "because"
        * (e.g. are the components entailed? presupposed?)
    
We did a replication and extension of @lucas2015improved's counterfactual experiments. We asked people to rate explanations, e.g. "A was present because B was present."

```{r load_data}
explanantes = c(story1="A", story2="B", story3="A",
                story4="C", story5="C", story6="B")
tags = c(
  story1A="match", story1B="miss",
  story2A="match", story2B="match", story2C="miss",
  story3A="unclear", story3B="match", story3C="unclear",
  story4A="match", story4B="match", story4C="miss", story4D="miss",
  story5A="match", story5B="match", story5C="match", story5D="miss",
  story6A="match", story6B="match", story6C="miss"
)

df = read.csv("../../data/2015-exp4-and-exp5-lkrep.csv") %>%
  select(workerid, story_index, story, variable, rating, version) %>%
  mutate(source = "replication human",
         workerid = as.character(workerid),
         variable = ifelse(story=="story3a" & variable=="B",
                           "C", as.character(variable))) %>%
  filter(version=="explanation")

expt.aggr = df %>%
  mutate(story = ifelse(story %in% c("story3a", "story3b"),
                        "story3", as.character(story))) %>%
  group_by(story, variable) %>%
  summarise(
    low = ci.low(rating),
    high = ci.high(rating),
    explanans = explanantes[story[1]],
    rating = mean(rating),
    tag = tags[paste(story, variable, sep="")[1]]) %>%
  mutate(source="human")
```

```{r plot_experiment}
expt.aggr %>%
  ggplot(., aes(x=variable, y=rating, colour=source, group=source)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(x=variable, ymin=low, ymax=high), width=0) +
  facet_wrap(~story, scale="free") +
  ylim(0, 1)
```

We implement a model (CF.DIFF) where the speaker/explainer chooses the explanans that maximizes the counterfactual *change* in the explanandum.

The items where this model diverges from people's intutions *seem* to be ones in which the explanation communicates something about the causal **structure** (e.g. B is or is not directly causally downstream of A).

```{r run_cf.diff}
lk = function(num, variable, model) {
  message(paste("story", num))
  explanations.model.rs = webppl(
    program_file=paste(model, "/lk", num, "/autoexpanded.wppl", sep=""),
    inference_opts = list(method="enumerate"),
    model_var = paste("explain('", variable, "')", "", sep=""),
    packages = "./node_modules/jsUtils")
  return(explanations.model.rs %>%
           mutate(story = paste("story", num, sep=""),
                  variable = variable,
                  source = model,
                  tag = tags[paste(story, variable, sep="")]) %>%
           rename(explanans = support, rating = prob) %>%
           mutate(low = rating, high = rating))
}
cf.diff = do.call(rbind, list(
  lk(1, "A", "cf.diff"), lk(1, "B", "cf.diff"), #fast enough
  lk(2, "A", "cf.diff"), lk(2, "B", "cf.diff"), lk(2, "C", "cf.diff"),
  #hella slow with flip(A ? 0.75 : 0) but fine with A?flip(0.75):false
  lk(3, "A", "cf.diff"), lk(3, "B", "cf.diff"), lk(3, "C", "cf.diff"),  #fast enough
  lk(4, "A", "cf.diff"), lk(4, "B", "cf.diff"), lk(4, "C", "cf.diff"), lk(4, "D", "cf.diff"),  #fast enough
  lk(5, "A", "cf.diff"), lk(5, "B", "cf.diff"), lk(5, "C", "cf.diff"), lk(5, "D", "cf.diff"),  #fast enough
  lk(6, "A", "cf.diff"), lk(6, "B", "cf.diff"), lk(6, "C", "cf.diff") #dunno
)) %>% 
  filter((story=="story1" & explanans==explanantes["story1"]) |
           (story=="story2" & explanans==explanantes["story2"]) |
           (story=="story3" & explanans==explanantes["story3"]) |
           (story=="story4" & explanans==explanantes["story4"]) |
           (story=="story5" & explanans==explanantes["story5"]) |
           (story=="story6" & explanans==explanantes["story6"]))
```

```{r plot_cf.diff}
rbind(cf.diff, expt.aggr) %>%
  gather("statistic", "value", c(rating, low, high)) %>%
  mutate(statistic = paste(source, statistic, sep=".")) %>%
  select(-source) %>%
  spread(statistic, value) %>%
  ggplot(., aes(x=cf.diff.rating, y=human.rating, colour=tag)) +
  geom_point() +
  geom_errorbar(aes(ymin=human.low, ymax=human.high), width=0) +
  geom_errorbarh(aes(xmin=cf.diff.low, xmax=cf.diff.high), height=0) +
  ylim(0,1) +
  xlim(0,1) +
  scale_colour_manual(values=c("green3", "deeppink3", "yellow3")) +
  geom_abline(intercept = 0, slope=1, colour="gray") +
  ggtitle("CF.DIFF")
ggsave("cf.diff.png", width=6, height=4)
```

We implement a second model (RSA.BASE) where there are no structural parameters, there *are* alternative possible inputs, the speaker knows all of the actual input and output values, and the listener knows none of them.
  

```{r}
lk = function(num, model) {
  message(paste("story", num))
  message(paste(model, "/lk", num, "/autoexpanded.wppl", sep=""))
  explanations.model.rs = webppl(
    program_file=paste(model, "/lk", num, "/autoexpanded.wppl", sep=""),
    inference_opts = list(method="enumerate"),
    model_var = "speaker",
    packages = "./node_modules/jsUtils")
  return(explanations.model.rs %>%
           mutate(story = paste("story", num, sep=""),
                  source = model))
}
rsa.base.full.data = do.call(rbind, list(
  lk(1, "rsa.base"),
  lk(2, "rsa.base"),
  lk(3, "rsa.base"),
  lk(4, "rsa.base"),
  lk(5, "rsa.base"),
  lk(6, "rsa.base")
))
```

```{r}
rsa.base = data.frame(
  story=c("story1", "story1",
          "story2", "story2", "story2",
          "story3", "story3", "story3",
          "story4", "story4", "story4", "story4",
          "story5", "story5", "story5", "story5",
          "story6", "story6", "story6"),
  variable=c("A", "B",
             "A", "B", "C",
             "A", "B", "C",
             "A", "B", "C", "D",
             "A", "B", "C", "D",
             "A", "B", "C"),
  explanans=c("A", "A",
              "B", "B", "B",
              "A", "A", "A",
              "C", "C", "C", "C",
              "C", "C", "C", "C",
              "B", "B", "B"),
  explanation=c(
    "A because A", "B because A",
    "A because B", "B because B", "C because B",
    "A because A", "B because A", "! C because A",
    "! A because ! C", "! B because ! C", "! C because ! C", "! D because ! C",
    "A because C", "B because C", "C because C", "D because C",
    "! A because ! B", "! B because ! B", "C because ! B")
) %>%
  mutate(rating = sapply(as.character(explanation), function(expl) {
    return(sum((rsa.base.full.data %>% filter(as.character(support)==expl))$prob))
  }),
  ### FIX THIS TO DO TODO
  ## this should be rescaled within story.
  ## otherwise I'm rendering the data meaningless
  # rating = rating / max(rating),
  low = rating,
  high = rating,
  source = "rsa.base",
  tag = tags[paste(story, variable, sep="")]) %>%
  select(story, variable, low, high,
         explanans, rating, tag, source)
```

```{r plot_rsa.base}
rbind(rsa.base, expt.aggr) %>%
  gather("statistic", "value", c(rating, low, high)) %>%
  mutate(statistic = paste(source, statistic, sep=".")) %>%
  select(-source) %>%
  spread(statistic, value) %>%
  ggplot(., aes(x=rsa.base.rating, y=human.rating, colour=tag)) +
  geom_point() +
  geom_errorbar(aes(ymin=human.low, ymax=human.high), width=0) +
  geom_errorbarh(aes(xmin=rsa.base.low, xmax=rsa.base.high), height=0) +
  ylim(0,1) +
  # xlim(0,1) +
  scale_colour_manual(values=c("green3", "deeppink3", "yellow3")) +
  geom_abline(intercept = 0, slope=1, colour="gray") +
  ggtitle("RSA.BASE")
ggsave("rsa.base.png", width=6, height=4)
```

# References