---
title: "Explanations with 2 causes"
author: "Erin Bennett"
output: html_document
---

```{r global_options, include=FALSE}
rm(list=ls())
library(knitr)
knitr::opts_chunk$set(
  echo=F, warning=F, #cache=T, 
  message=F, #sanitiz =T, 
  fig.width = 5, fig.height = 3)
```

```{r}
source("~/Settings/startup.R")
```

```{r yak_shaving_for_plot_dodge}
my_dodge <- function(n_colors=2, width = NULL, preserve = c("total", "single")) {
  ggproto(NULL, MyDodge,
          width = width,
          n_colors = n_colors,
          preserve = match.arg(preserve)
  )
}

# Detect and prevent collisions.
# Powers dodging, stacking and filling.
my_collide <- function(data, width = NULL, name, strategy, ..., check.width = TRUE, reverse = FALSE) {
  # Determine width
  if (!is.null(width)) {
    # Width set manually
    if (!(all(c("xmin", "xmax") %in% names(data)))) {
      data$xmin <- data$x - width / 2
      data$xmax <- data$x + width / 2
    }
  } else {
    if (!(all(c("xmin", "xmax") %in% names(data)))) {
      data$xmin <- data$x
      data$xmax <- data$x
    }
    
    # Width determined from data, must be floating point constant
    widths <- unique(data$xmax - data$xmin)
    widths <- widths[!is.na(widths)]
    
    #   # Suppress warning message since it's not reliable
    #     if (!zero_range(range(widths))) {
    #       warning(name, " requires constant width: output may be incorrect",
    #         call. = FALSE)
    #     }
    width <- widths[1]
  }
  
  # Reorder by x position, then on group. The default stacking order reverses
  # the group in order to match the legend order.
  if (reverse) {
    data <- data[order(data$xmin, data$group), ]
  } else {
    data <- data[order(data$xmin, -data$group), ]
  }
  
  
  # Check for overlap
  intervals <- as.numeric(t(unique(data[c("xmin", "xmax")])))
  intervals <- intervals[!is.na(intervals)]
  
  if (length(unique(intervals)) > 1 & any(diff(scale(intervals)) < -1e-6)) {
    warning(name, " requires non-overlapping x intervals", call. = FALSE)
    # This is where the algorithm from [L. Wilkinson. Dot plots.
    # The American Statistician, 1999.] should be used
  }
  
  if (!is.null(data$ymax)) {
    plyr::ddply(data, "xmin", strategy, ..., width = width)
  } else if (!is.null(data$y)) {
    data$ymax <- data$y
    data <- plyr::ddply(data, "xmin", strategy, ..., width = width)
    data$y <- data$ymax
    data
  } else {
    stop("Neither y nor ymax defined")
  }
}

MyDodge <- ggproto(
  "MyDodge",
  Position,
  required_aes = "x",
  width = NULL,
  n_colors = 2,
  preserve = "total",
  setup_params = function(self, data) {
    # if (is.null(data$xmin) && is.null(data$xmax) && is.null(self$width)) {
    #   warning("Width not defined. Set with `position_dodge(width = ?)`",
    #           call. = FALSE)
    # }
    
    if (identical(self$preserve, "total")) {
      n <- NULL
    } else {
      n <- max(table(data$xmin))
    }
    
    list(
      width = self$width,
      n = n,
      n_colors = self$n_colors
    )
  },
  
  compute_panel = function(data, params, scales) {
    my_collide(
      data,
      params$width,
      name = "my_dodge",
      strategy = my_dodge_strategy(params$n_colors),
      n = params$n,
      check.width = FALSE
    )
  }
)

# Dodge overlapping interval.
# Assumes that each set has the same horizontal position.
my_dodge_strategy <- function(n_colors = 2) {
  return(function(df, width, n = NULL) {
    if (is.null(n)) {
      n <- length(unique(df$group))
    }
    
    if (n == 1)
      return(df)
    
    if (!all(c("xmin", "xmax") %in% names(df))) {
      df$xmin <- df$x
      df$xmax <- df$x
    }
    
    d_width <- max(df$xmax - df$xmin)
    
    # Have a new group index from 1 to number of groups.
    # This might be needed if the group numbers in this set don't include all of 1:n
    groupidx <- match(df$group, sort(unique(df$group)))
    
    # width is 1 and d_width is 1
    # print(df$x) # do not change df$x. that's just the value of A
    if (n%%2==1) {
      n=n+1
      # groupidx = ifelse(groupidx==1, groupidx, groupidx+1)
    }
    
    n = n/n_colors
    groupidx = n - (groupidx %% n)
    
    # print(n)
    # print(groupidx)
    # print(df$x)
    # print(width)
    # print(d_width)
    
    # Find the center for each group, then use that to calculate xmin and xmax
    df$x <- df$x + width * ((groupidx - 0.5) / n - .5)
    df$xmin <- df$x - d_width / n / 2
    df$xmax <- df$x + d_width / n / 2
    
    df
  })
}
```

```{r}
forward_sample = webppl(
  program_file = "model.wppl",
  model_var = "forward_sample()",
  inference_opts = list(method="enumerate")
)

actual = forward_sample %>%
  rename(A = actual.A, B=actual.B, E=actual.E,
         AE = causal_parameters.AE, BE=causal_parameters.BE) %>%
  group_by(A, B, E, AE, BE) %>%
  summarise(prob=sum(prob)) %>%
  as.data.frame
```

The listener has uncertainty about whether each of two possible causes A and B are actually causally linked to the effect E.

At first, let's assume the background probability of E is 0, i.e. there are no other causes of E. That means that when neither A nor B are linked to E, E can only ever be false.

We set causal strength to 0.9.

```{r}
actual %>% 
  mutate(AE = paste("A->E:", AE),
         BE = paste("B->E:", BE)) %>%
  ggplot(., aes(x=B, y=A, fill=E, alpha=prob)) +
  facet_grid(AE ~ BE) +
  geom_tile(position=my_dodge(n_colors=2, width=1)) +
  geom_text(aes(label=round(prob, 2)), alpha=1, position = my_dodge(n_colors=2, width=1), size=3) +
  scale_fill_brewer(type = "qual", palette = 6)
```

```{r, fig.width=12, fig.height=8}
# blue = "#377eb8"
# red = "#e41a1c"
# pink = "#ed12a4"
# sky = "#13ceef"
# orange = "#e07802"
# forward_sample %>%
#   mutate(A = paste(actual.A, counterfactual.A),
#          B = paste(actual.B, counterfactual.B),
#          E = paste(actual.E, counterfactual.E),
#          AE = paste("A->E:", causal_parameters.AE),
#          BE = paste("B->E:", causal_parameters.BE)) %>%
#   ggplot(., aes(x=A, y=B, fill=E, alpha=prob)) +
#   facet_grid(AE ~ BE) +
#   geom_tile(position=my_dodge(n_colors=4), height=0.99, width=0.98) +
#   geom_text(aes(label=round(prob, 2)), alpha=1,
#             position = my_dodge(width=1, n_colors=4), size=2.5) +
#   geom_hline(yintercept = 2.5) +
#   geom_vline(xintercept = 2.5) +
#   # scale_fill_brewer(type="qual", palette = 6)
#   scale_fill_manual(values=c(red, orange, sky, blue))
```

Suppose we condition on the background knowledge that all of the variables (A, B, and E) were true in the actual world.

This means that one of the links (A->E or B->E) must be there.

Here are the counterfactual values for A, B, and E:

```{r}
background_knowledge = "{A: true, B: true, E: true}"
forward_sample_with_background_knowledge = webppl(
  program_file = "model.wppl",
  model_var = paste("forward_sample(", background_knowledge,")", sep=""),
  inference_opts = list(method="enumerate")
) %>%
  mutate(A = counterfactual.A,
         B = counterfactual.B,
         E = counterfactual.E,
         AE = paste("A->E:", causal_parameters.AE),
         BE = paste("B->E:", causal_parameters.BE))
```

```{r}
forward_sample_with_background_knowledge %>%
  ggplot(., aes(x=B, y=A, fill=E, alpha=prob)) +
  facet_grid(AE ~ BE) +
  geom_tile(position=my_dodge(n_colors=2, width=1), height=0.99, width=0.98) +
  geom_text(aes(label=round(prob, 2)), alpha=1, position = my_dodge(n_colors=2, width=1), size=3) +
  scale_fill_brewer(type = "qual", palette = 6)
```

Now if we want to evaluate "if A_CF then E_CF", we rule out the situation where A_CF is true and E_CF is false. This rules out an extremely low probabiity event.

```{r}
forward_sample_with_background_knowledge %>%
  filter(!A | E) %>%
  mutate(prob = prob/sum(prob)) %>%
  ggplot(., aes(x=B, y=A, fill=E, alpha=prob)) +
  facet_grid(AE ~ BE) +
  geom_tile(position=my_dodge(n_colors=2, width=1)) +
  geom_text(aes(label=round(prob, 2)), alpha=1, position = my_dodge(n_colors=2, width=1), size=3) +
  scale_fill_brewer(type = "qual", palette = 6)
```

Given just the background knowledge, the more causes are enabled, the more likely a set of causal parameters is.

```{r}
forward_sample_with_background_knowledge %>%
  group_by(AE, BE) %>% summarise(prob = sum(prob)) %>%
  as.data.frame %>%
  ggplot(., aes(x=BE, y=AE, alpha=prob)) +
  geom_tile(fill="gray") +
  geom_text(aes(label=round(prob,2)), alpha=1)
```

After conditioning on "if A_CF then E_CF", it becomes much more likely that there is a causal link between A and E.

```{r}
forward_sample_with_background_knowledge %>%
  filter(!A | E) %>%
  mutate(prob = prob / sum(prob)) %>%
  group_by(AE, BE) %>% summarise(prob = sum(prob)) %>%
  as.data.frame %>%
  ggplot(., aes(x=BE, y=AE, alpha=prob)) +
  geom_tile(fill="gray") +
  geom_text(aes(label=round(prob,2)), alpha=1)
```
